Geometry of Ranked Nearest Neighbour Interchange Space of Phylogenetic Trees

### 2019-10-29 paper.tex
proof lemma 1

\todo{Add a bit more space in all figures at the bottom to have at least a standard interval above he caption.}
\todo{In this figure -- rm arrows (think how to place the RNNI edges), possibly avoid using colour, and highlight where the rank move is made (hollow nodes?)}
\todo{Two of these trees are caterpillar trees.
Do you think it would be helpful to draw them as caterpillars to prepare readers for the rest of the paper?}
\begin{figure}[H]
	\centering
	\includegraphics[width=0.8\textwidth]{RNNI}
    \vspace{2pt}
	\caption{Two types of operation that define edges of the $\rnni$ graph.
    The move at the top left is a rank move while the trees in the triangle are connected by $\nni$ moves.}
	\label{fig:RNNI}
\end{figure}

@@

\begin{lemma}
    Let $T$ and $R$ be two trees on which we apply $\findpath$ to compute a path from $T$ to $R$.
    We assume that $\hat T \neq T,R$ is a tree on the path computed by this algorithm, which needs to be updated in Line~\ref{alg:findpath:line:move_set_down} of Algorithm~\ref{alg:find_path}.
    Then there is always exactly one $\rnni$ move that can update $\hat T$ to a tree $T'$ such that $\mrca_{T'}(C_i) < \mrca_{\hat T}(C_i)$ for cluster $C_i$ as defined in Line~\ref{alg:find_path:line:cluster}.
    \label{lemma:mrca_move}
\end{lemma}

@@

    Let $T$ and $R$ be trees and $[C_1, \ldots, C_n]$ the cluster representation
    \todo{LC: How do we call this representation? Introduce it in the beginning of alg chapter!}
    of $R$ as introduced above.
    Let $\hat T$ be a tree as in Line~\ref{alg:findpath:line:move_set_down} of Algorithm~\ref{alg:find_path} and $C_i$ the cluster defined in Line~\ref{alg:find_path:line:cluster}.
    We want to prove that there is exactly one $\rnni$ move possible on $\hat T$ that results in a tree $T'$ where $\mrca_{T'}(C_i) < \mrca_{\hat T}(C_i)$ if $C_i$ is not induced by a node in $\hat T$ already.

    To do this, we distinguish the following two cases:
    Either there is an edge between the node $v = \mrca_{\hat T}(C_i)$ and the node $w$ with rank $rank_{\hat T}(w) = rank_{hat T}(v) - 1$, or there is no such edge.

    In the latter case, there is only one $\rnni$ move that could increase the rank of $v$.
    That is a rank move swapping the ranks of $v$ and $w$.

    If there is an edge $(v,w)$ in $\hat T$, there are two $\nni$ moves possible on this edge, which could possibly change the rank of the most recent common ancestor of $C_i$.
    We are now going to explain why only one of these two $\nni$ moves changes the rank of this most recent common ancestor.
    Therefore it is important to see that the clusters $C_1, \ldots, C_{i-1}$ already appear in $\hat T$ as they do in $R$, which is a direct result from the algorithm.
    It is also straight forward to see
    \todo{LC: is it?}
    that the $C_i$ either is the disjoint union $C_i = C_j \dot\cup C_k$ for $j,k < i$ or $C_i = C_j \dot\cup \{x\}$ for a taxon $x \in X \setminus (C_1 \cup \ldots \cup C_{i-1})$.

    Le us at first assume that $C_i = C_j \dot\cup C_k$ for some $j,k < i$.
    The tree $\hat T$ looks like the tree depicted in Figure~\ref{fig:mrca_move} where $T_w^1$ and $T_w^2$ are the two subtrees rooted in the children of $w$ and $T_v$ is the subtree whose root is child of $v$ and does not include $w$.
    Because it is $v = \mrca_{\hat T}(C_i)$, either $C_j$ or $C_k$ is a subset of the taxon set of $T_v$.
    Without loss of generality we assume that it is $C_j$.
    Since $C_k$ is already induced by a node in $\hat T$, all taxa in $C_k$ are either in $T_w^1$ or in $T_w^2$.
    If there were taxa of $C_k$ in both subtrees $T_w^1$ and $T_w^2$, $w$ would be the node inducing $C_k$ and hence it would be $k = i-1$, which means that $v$ induces $C_i$.
    Therefore, we know that $C_j$ is a subset of the taxon set of $T_v$ and $C_k$ is subset of the taxon set of one of the subtrees $T_w^1$ or $T_w^2$.
    In Figure~\ref{fig:mrca_move} one can see the two moves possible on $(v,w)$.
    Depending on which of the subtrees $T_w^1$ or $T_w^2$ includes the taxa of $C_k$, one can see that only one of the two $\nni$ moves results in a tree where the tank of the most recent common ancestor of $C_i$ is below its rank in $\hat T$.

    Now we consider the case that $C_i = C_j \dot\cup \{x\}$ for a taxon $x \in X \setminus (C_1 \cup \ldots \cup C_{i-1})$.
    For the same reason as above, $C_j$ is in subset of the taxon set of one of the subtrees $T_w^1, T_w^2$ or $T_v$ of $\hat T$ as depicted in Figure~\ref{fig:mrca_move}.
    Similarly $x$ is taxon of one of these three subtrees in a way that either $C_j$ or $x$ is included in $T_v$, because $v = \mrca_{\hat T}(C_i)$.
    So there is just one of the two subtrees $T_w^1$ and $T_w^2$ that neither includes $C_j$ nor $x$.
    Therefore, the $\nni$ move that moves this subtree to the position where $T_v$ has been before, is the only $\nni$ move that decreases the rank of $\mrca_{\hat T}(C_i)$.

    We can conclude that there is in any case exactly one $\rnni$ move that can be performed on $\hat T$ that decreases the rank of $\mrca_{\hat T}(C_i)$ in Line~\ref{alg:findpath:line:move_set_down} of Algorithm~\ref{alg:find_path}.

    \begin{figure}[H]
    	\centering
    	\includegraphics[width=0.5\textwidth]{mrca_move}
        \vspace{2pt}
    	\caption{\todo{caption}}
    	\label{fig:mrca_move}
    \end{figure}

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=\textwidth]{NNI_vs_RNNI}
    \vspace{2pt}
	\caption{Paths between caterpillar trees $T$ and $R$: The solid paths are paths in $\rnni$, the dashed one is a path that is only possible in the plain $\nni$ graph.
    The upper path, which is the $\rnni$ path that is the natural extension of the shortest $\nni$ path, is longer than a shortest $\rnni$ path as the one at the bottom.}
	\label{fig:NNI_vs_RNNI}
\end{figure}

@@

    \todo{LC: replace this figure}
    \begin{figure}[H]
    	\centering
        \begin{tabular}{|c|c|c|c|c|c|c|c|c|c|c|c|}
            \hline
            $x_{k+1}$ & $x_k$ & $\ldots$ & $b$ & $a$ & \multicolumn{3}{c|}{$\ldots$} & $x_n$ & $x_{k-1}$ & $\ldots$ & $x_1$\\
            \hline
            \multicolumn{12}{c}{}\\
            \hline
            $x_{k+1}$ & $x_{k+2}$ & $\ldots$ & $b$ & $x_k$ & $\ldots$ & $a$ & $\ldots$ & $x_n$ & $x_{k-1}$ & $\ldots$ & $x_1$\\
            \hline
            \multicolumn{12}{c}{}\\
            \multicolumn{12}{c}{}\\
            \hline
            $x_{k+1}$ & $x_k$ & $\ldots$ & $a$ & $b$ & \multicolumn{3}{c|}{$\ldots$} & $x_n$ & $x_{k-1}$ & $\ldots$ & $x_1$\\
            \hline
            \multicolumn{12}{c}{}\\
            \hline
            $x_{k+1}$ & $x_{k+2}$ & $\ldots$ & $x_k$ & $b$ & $\ldots$ & $a$ & $\ldots$ & $x_n$ & $x_{k-1}$ & $\ldots$ & $x_1$\\
            \hline
        \end{tabular}
        \vspace{2pt}
    	\caption{Loop number $k$ of $\csort$ between $T$ and $R$ (in list representation) as described in proof of Theorem~\ref{thm:caterpillar_convex}.
        The trees are represented as sequences that are being sorted by $\csort$.
        Right before loop number $k$, taxa $x_1, \ldots, x_{k-1}$ are already at the same position as in $R$.
        At the top: sequences between $T$ and $R$ -- $x_k$ and $b$ swap in loop number $k$, but $x_k$ and $a$ not.
        At the bottom: sequences between $T'$ and $R$ -- $a$ and $b$ swap in loop number $k$; $x_k$ and $b$ will swap later}
    	\label{fig:cat_convex_seq_example}
    \end{figure}

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=\textwidth]{NNI_NP_proof}
    \vspace{2pt}
    \caption{Example of a shortest path between two trees $T$ and $R$ in $\nni$.
    All leaves labelled with dots are labelled by taxa in $\{1, \ldots, n\}$, squares represent taxa in $\{n+1, \ldots, 2n\}$.
    In \autocite{Li1996-zw} it is proven that there is such a labelling that ensures that there is no shortest path in $\nni$ where the clusters shared between $T$ and $R$ are preserved.
    Instead, the depicted path where $n$ cherries are built ad then sorted before being resolved is a shortest path}
	\label{fig:NNI_NP_proof}
\end{figure}

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=\textwidth]{splitthm_counterexample}
    \vspace{2pt}
	\caption{The split $123|45$ is present in $T$ and $R$, but the path at the top is a shortest path (computed by $\csort$) where none of the trees contains this split.
    However, this split is maintained on the path at the bottom, which is a shortest path as well.}
	\label{fig:splitthm_counterexample}
\end{figure}

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=0.7\textwidth]{findpath_not_symmetric}
    \vspace{2pt}
	\caption{Two paths computed by $\findpath$. At the top $\findpath(T,R)$, at the bottom $\findpath(R,T)$}
	\label{fig:findpath_not_symmetric}
\end{figure}

>>>>

\todo{Add a bit more space in all figures at the bottom to have at least a standard interval above he caption.}
\todo{In this figure -- rm arrows (think how to place the RNNI edges), possibly avoid using colour, and highlight where the rank move is made (hollow nodes?)}
\todo{Two of these trees are caterpillar trees.
Do you think it would be helpful to draw them as caterpillars to prepare readers for the rest of the paper?}
\begin{figure}[H]
	\centering
	\includegraphics[width=0.8\textwidth]{RNNI}
    \vspace{12pt}
	\caption{Two types of operation that define edges of the $\rnni$ graph.
    The move at the top left is a rank move while the trees in the triangle are connected by $\nni$ moves.}
	\label{fig:RNNI}
\end{figure}

@@

The following lemma shows that $\findpath$ gives a path between the two input trees, indeed, by proving that there is always an $\rnni$ move possible as descibred in Line~\ref{ald:findpath:line:move_set_down} of Algorithm~\ref{alg:find_path}.

\begin{lemma}
    Let $T$ and $R = [C_1, \ldots, C_{n-2}]$ be two trees and $p:= \findpath (T,R)$ a path from $T$ to $R$.
    Let $T' \neq T,R$ be a tree on $p$ such that $C_1, \ldots, C_{k-1}$ are induced by internal nodes of $T'$, but $C_k$ is not.
    Then there is always exactly one $\rnni$ move on $T'$ that results in a tree $T''$ such that $\mrca_{T''}(C_i) < \mrca_{T'}(C_i)$.
    \label{lemma:mrca_move}
\end{lemma}

@@

    Let $T, R$ and $T'$ be trees as defined in the lemma and $[C_1, \ldots, C_{n-2}]$ the cluster representation of $R$.
    We want to prove that there is exactly one $\rnni$ move on $T'$ that results in a tree $T''$ where $\mrca_{T''}(C_k) < \mrca_{T'}(C_k)$.
    Notice, that the only $\rnni$ move on $T$ that could decrease the rank of $\mrca_{T'}(C_k)$ are $\nni$ movs on the edge between $\mrca_{T'}(C_k)$ and its child (if this edge has length one), or a rank move between $\mrca_{T'}(C_k)$ and the node with rank one less.
    Therefore, we will now distinguish these two cases:
    Either there is an edge $(v,w)$ between nodes $v = \mrca_{T'}(C_k)$ and $w$ with rank $rank_{T'}(w) = rank_{T'}(v) - 1$, or there is no such edge.

    In the latter case the only $\rnni$ move that could increase the rank of $v$ is a rank move swapping the ranks of $v$ and $w$.

    If there is an edge $(v,w)$ in $T'$ there are two $\nni$ moves possible on this edge.
    We are now going to explain why only one of these two $\nni$ moves changes the rank of this most recent common ancestor.
    By definition we know that $C_1, \ldots, C_{k-1}$ are induced by the nodes of rank $1, \ldots, k-1$, respectively, in $T'$ already.
    It is important to notice that $C_k$ either is the disjoint union $C_k = C_i \dot\cup C_j$ for $i,j < k$ or $C_k = C_i \dot\cup \{x\}$ for a taxon $x \in X \setminus (C_1 \cup \ldots \cup C_{k-1})$.

    Le us at first assume that $C_k = C_i \dot\cup C_j$ for some $i,j < k$.
    The tree $T'$ looks like the tree depicted in the left of Figure~\ref{fig:mrca_move} where $T_w^1$ and $T_w^2$ are the two subtrees rooted in the children of $w$ and $T_v$ is the subtree whose root is child of $v$ and does not include $w$.
    Since we know that all clusters $C_1,\ ldots, C_{k-1}$ are induced by some nodes in $T'$, each of these clusters must be a subset of the taxon set of one of the three subtrees $T_w^1,T_w^2$, or $T_v$.
    In particular, there can't be elements of the same cluster $C_m$ in more than one of these subtrees.
    This and the fact that $v = \mrca_{T'}(C_i \dot\cup C_j)$ yields that either $C_i$ or $C_j$ is subset of the taxon set of $T_v$ and the other cluster is subset of the taxon set of either $T_w^1$ or $T_w^2$.

    Without loss of generality we assume that $C_i$ is included in $T_v$.
    Otherwise, change the indices $i$ and $j$.
    Then $C_i$ is a subset of the taxon set of either $T_w^1$ or $T_w^2$.
    Therefore, just one of the two $\nni$ moves on $(v,w)$, as depicted in Figure~\ref{fig:mrca_move}, moves the most recent common ancestor of $C_k = C_i \dot \cup C_j$ down.

    Now we consider the case that $C_k = C_i \dot\cup \{x\}$ for a taxon $x \in X \setminus (C_1 \cup \ldots \cup C_{k-1})$.
    For the same reason as above, $C_i$ is in subset of the taxon set of one of the subtrees $T_w^1, T_w^2$ or $T_v$ of $\hat T$ as depicted in Figure~\ref{fig:mrca_move}.
    Similarly $x$ is taxon of one of these three subtrees in a way that either $C_i$ or $x$ is included in $T_v$, because $v = \mrca_{T'}(C_i \dot \cup \{x\})$.
    So there is just one of the two subtrees $T_w^1$ and $T_w^2$ that neither includes $C_j$ nor $x$.
    As in the previous case, just one of the two $\nni$ moves on $(v,w)$ as illustrated in Figure~\ref{fig:mrca_move} decreases the rank of $\mrca_{T'}(C_i \dot \cup \{x\})$.

    We can conclude that in any case there is exactly one $\rnni$ move possible on $T'$ that decreases the rank of $\mrca_{T'}(C_k)$ in Line~\ref{alg:findpath:line:move_set_down} of Algorithm~\ref{alg:find_path}.

    \begin{figure}[H]
    	\centering
    	\includegraphics[width=0.5\textwidth]{mrca_move}
        \vspace{12pt}
        \caption{Two $\nni$ moves that are possible on the edge $(v,w)$ of the tree $T'$.
        They result in the trees on the right as described in the proof of Lemma~\ref{lemma:mrca_move}.}
    	\label{fig:mrca_move}
    \end{figure}

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=\textwidth]{NNI_vs_RNNI}
    \vspace{12pt}
	\caption{Paths between caterpillar trees $T$ and $R$: The solid paths are paths in $\rnni$, the dashed one is a path that is only possible in the plain $\nni$ graph.
    The upper path, which is the $\rnni$ path that is the natural extension of the shortest $\nni$ path, is longer than a shortest $\rnni$ path as the one at the bottom.}
	\label{fig:NNI_vs_RNNI}
\end{figure}

@@

    \todo{LC: replace this figure}
    \begin{figure}[H]
    	\centering
        \begin{tabular}{|c|c|c|c|c|c|c|c|c|c|c|c|}
            \hline
            $x_{k+1}$ & $x_k$ & $\ldots$ & $b$ & $a$ & \multicolumn{3}{c|}{$\ldots$} & $x_n$ & $x_{k-1}$ & $\ldots$ & $x_1$\\
            \hline
            \multicolumn{12}{c}{}\\
            \hline
            $x_{k+1}$ & $x_{k+2}$ & $\ldots$ & $b$ & $x_k$ & $\ldots$ & $a$ & $\ldots$ & $x_n$ & $x_{k-1}$ & $\ldots$ & $x_1$\\
            \hline
            \multicolumn{12}{c}{}\\
            \multicolumn{12}{c}{}\\
            \hline
            $x_{k+1}$ & $x_k$ & $\ldots$ & $a$ & $b$ & \multicolumn{3}{c|}{$\ldots$} & $x_n$ & $x_{k-1}$ & $\ldots$ & $x_1$\\
            \hline
            \multicolumn{12}{c}{}\\
            \hline
            $x_{k+1}$ & $x_{k+2}$ & $\ldots$ & $x_k$ & $b$ & $\ldots$ & $a$ & $\ldots$ & $x_n$ & $x_{k-1}$ & $\ldots$ & $x_1$\\
            \hline
        \end{tabular}
        \vspace{12pt}
    	\caption{Loop number $k$ of $\csort$ between $T$ and $R$ (in list representation) as described in proof of Theorem~\ref{thm:caterpillar_convex}.
        The trees are represented as sequences that are being sorted by $\csort$.
        Right before loop number $k$, taxa $x_1, \ldots, x_{k-1}$ are already at the same position as in $R$.
        At the top: sequences between $T$ and $R$ -- $x_k$ and $b$ swap in loop number $k$, but $x_k$ and $a$ not.
        At the bottom: sequences between $T'$ and $R$ -- $a$ and $b$ swap in loop number $k$; $x_k$ and $b$ will swap later}
    	\label{fig:cat_convex_seq_example}
    \end{figure}

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=\textwidth]{NNI_NP_proof}
    \vspace{12pt}
    \caption{Example of a shortest path between two trees $T$ and $R$ in $\nni$.
    All leaves labelled with dots are labelled by taxa in $\{1, \ldots, n\}$, squares represent taxa in $\{n+1, \ldots, 2n\}$.
    In \autocite{Li1996-zw} it is proven that there is such a labelling that ensures that there is no shortest path in $\nni$ where the clusters shared between $T$ and $R$ are preserved.
    Instead, the depicted path where $n$ cherries are built ad then sorted before being resolved is a shortest path}
	\label{fig:NNI_NP_proof}
\end{figure}

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=\textwidth]{splitthm_counterexample}
    \vspace{12pt}
	\caption{The split $123|45$ is present in $T$ and $R$, but the path at the top is a shortest path (computed by $\csort$) where none of the trees contains this split.
    However, this split is maintained on the path at the bottom, which is a shortest path as well.}
	\label{fig:splitthm_counterexample}
\end{figure}

@@

\begin{figure}[H]
	\centering
	\includegraphics[width=0.7\textwidth]{findpath_not_symmetric}
    \vspace{12pt}
	\caption{Two paths computed by $\findpath$. At the top $\findpath(T,R)$, at the bottom $\findpath(R,T)$}
	\label{fig:findpath_not_symmetric}
\end{figure}
